---
title: "Untitled"
format: pdf
editor: visual
---

# Code for Class Slides

```{r, include=FALSE}
library(tidyverse)
library(pokemon)
library(broom)

set.seed(1)
pokemon_sample <- pokemon |> slice_sample(prop=0.5)
write_csv(pokemon_sample, "pokemon_sample.csv")
```

### Example 1a: Deterministic function

Suppose we have a function of $x$: $f(x) = 3x + 5$. Then, even if we only have a few data points, we can find a new data point, as shown below for $f(1) = 8$.

```{r}
example_df <- tibble(x=c(0, 2, 5, 6),                      
                     f_x=c(5, 11, 20, 23))  
example_df |>   
  ggplot()+   
  geom_point(aes(x=x, y=f_x))+   
  geom_line(aes(x=x, y=f_x))+   
  geom_vline(xintercept=1, color="red", linetype="dashed")+   
  geom_hline(yintercept=8, color="purple", linetype="dashed")+   
  annotate("text", x=0.5, y=9, label= "f(1)=8", color="purple")+   
  labs(x="x", y="f(x)")+   
  theme_bw() 
```

### Example 1b: Adding Random Error

A slightly different version of the problem is if we have \$f(x) = 3x + 5 + \epsilon\$, where $\epsilon$ is a random error term. In this case, we use `rnorm()` to sample random normally-distributed random values when creating these data. Now, we cannot deterministically calculate $f(1)$, though the guess of $f(1) = 8$ does not seem unreasonable.

```{r}
set.seed(260) 
random_df <- tibble(x=c(0, 2, 5, 6),                     
                    f_x=c(5, 11, 20, 23)+rnorm(4),                     
                    f_x_true=c(5, 11, 20, 23))  
random_df |>   
  ggplot()+   
  geom_point(aes(x=x, y=f_x))+   
  geom_line(aes(x=x, y=f_x))+   
  geom_abline(slope=3, intercept=5, linetype="dashed")+   
  geom_vline(xintercept=1, color="red", linetype="dashed")+   
  geom_hline(yintercept=8, color="purple", linetype="dashed")+   
  annotate("text", x=0.5, y=9, label= "f(1)=8?", color="purple")+   
  labs(x="x", y="f(x)")+   
  theme_bw() 
```

### Example 1c: Multiple Data Points at same $x$ value

A further case is what if we don't just have 1 $f(x)$ for any $x$, but we have data pairs $(x_i, y_i)$ where $i$ is an index from $1$ to $n$. Here, there could be multiple observation at a particular fixed $x$. We now don't think of any function $f$ as a fixed function that dictates $f(x)$, but instead as a **model** for explaining the relationship between $x$ and $y$. Here, we generate $(x, y)$ pairs from a true model of $f(x) = 3x + 5$, where we add random normally-distributed error to both $x$ and $y$.

```{r}
set.seed(260)
random_df_10pts <- tibble(x = rep(c(0,2,5,6), 10)+rnorm(40, sd=0.5),
                          y = rep(c(5, 11, 20, 23), 10)+rnorm(40),
                          f_x_true=rep(c(5, 11, 20, 23), 10))

random_df_10pts |>   
  ggplot()+   
  geom_point(aes(x=x, y=y))+   
  labs(x="x", y="y")+   
  theme_bw() 
```

Let's once again plot the true model between the $x$ and $y$ variables in the scatterplot:

```{r}
random_df_10pts |>   
  ggplot()+   
  geom_point(aes(x=x, y=y))+  
  geom_abline(slope=3, intercept=5, linetype="dashed")+   
  geom_vline(xintercept=1, color="red", linetype="dashed")+   
  geom_hline(yintercept=8, color="purple", linetype="dashed")+   
  annotate("text", x=0.5, y=9, label= "f(1)=8?", color="purple")+   
  labs(x="x", y="y")+   
  theme_bw() 
```

What does the error look like? We can visualize the error terms using `geom_segment()`:

```{r}
random_df_10pts |>   
  ggplot()+   
  geom_point(aes(x=x, y=y))+ 
  geom_abline(slope=3, intercept=5, linetype="dashed")+  
  geom_segment(aes(x = x, y = y, xend = x, yend = 3*x + 5), color = "red") +  # Vertical lines
  labs(x="x", y="y")+   
  theme_bw() 
```

```{r, eval=FALSE, echo=FALSE}
set.seed(1)
pop_n <- 200
dat1 <- tibble(x = runif(n=pop_n, min=5, max=20), 
               y = 2*x + 3 + rnorm(n=pop_n, mean=0, sd = 2)) 

dat1 |>
  ggplot(aes(x=x, y=y)) +
  geom_point(alpha=1) +
  geom_smooth(method="lm", se=FALSE, lty="dashed", lwd=1, color="red") +
  annotate("text", x = Inf, y = 30, 
           label = expression(y == beta[0] + beta[1]*x + epsilon), 
           size = 5, hjust = 1, vjust=1, color="red") +
  labs(title = paste("Population of size 200"))

dat1 |> 
  slice_sample(n=30) |>
  ggplot(aes(x=x, y=y)) +
  geom_point() +
  labs(title = paste("Sample of size n=30"))
```

### Example 2: Many different kinds of relationships!

Models don't exclusively have to be *linear* in form. We will likely want to specify a non-linear model for some of the data plotted below:

```{r}
set.seed(260)
random_df_2 <- tibble(x = runif(1000, -10, 10),
                      y1 = 5+3*x+rnorm(1000),
                      y2 = 5+3*x^2+rnorm(1000, sd=5),
                      y3 = abs(5+3*x)+rnorm(1000),
                      y4 = ifelse(5+3*x+rnorm(1000, sd=5)>12, 1, 0))

random_df_2 |>   
  ggplot()+   
  geom_point(aes(x=x, y=y1), alpha=0.5)+   
  labs(x="x", y="y")+
  theme_bw() 
```

```{r}
random_df_2 |>   
  ggplot()+   
  geom_point(aes(x=x, y=y2), alpha=0.5)+   
  labs(x="x", y="y")+   
  theme_bw() 
```

```{r}
random_df_2 |>   
  ggplot()+   
  geom_point(aes(x=x, y=y3), alpha=0.5)+   
  labs(x="x", y="y")+   
  theme_bw() 
```

```{r}
random_df_2 |>   
  ggplot()+   
  geom_point(aes(x=x, y=y4), alpha=0.5)+   
  labs(x="x", y="y")+   
  theme_bw()
```

## Shoeprints and Heights

```{r}
heights <- read_csv("heights.csv")
glimpse(heights)

# Scatterplot (no line)
heights |> ggplot(aes(x=shoePrint, y = height)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "blue") +
  xlab("Length of shoeprint (in cm)") +
  ylab("Height (in cm)") +
  theme_minimal()

# Adding the lines
model_heights <- lm(height ~ shoePrint, data=heights)
heights |> 
  mutate(fitted_values = fitted(model_heights)) |>
  ggplot(aes(x=shoePrint, y = height)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "blue") +
  # Add vertical lines from points to regression line
  geom_segment(aes(xend = shoePrint, yend = fitted_values), 
               color = "red", alpha = 0.7) +
  labs(x = "Length of shoeprint (in cm)", 
       y = "Height (in cm)")
```

## Pokemon

```{r}
pokemon_sample <- read_csv("pokemon_sample.csv")

psychic_pokemon <- pokemon_sample |> 
  filter((type_1 == "psychic") | (type_2 == "psychic"))

psychic_pokemon |>
  ggplot(aes(x = speed, y = special_attack)) +
  geom_point() + 
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    title = "Psychic Pokemon: Special-Attack vs Speed",
    x = "Speed Points",
    y = "Special-Attack Points")

# Fitting a linear model
model_1 <- lm(special_attack ~ speed, data = psychic_pokemon)
model_1 |> tidy()
beta1_hat <- model_1 |> 
   tidy() |>
   filter(term=="speed") |> 
   select(estimate) |>
   pull()
beta1_hat


# Fitting another linear model: hit points ~ base experience points
model_2 <- lm(hp ~ base_experience, data=pokemon_sample)
model_2 |> tidy()


```
