---
title: "STA130 Week 3 - Data Wrangling"
author: "Emily Somerset"
format: pdf
---

```{r libraries, include=FALSE}
library(tidyverse)
library(openintro)
library(ggplot2)
```

# Before this demo

1. Start by reviewing the lecture material.
2. Review the code in `week3-supplementary-materials.qmd`.
3. Try these questions on your own!
4. Watch the demo video for some additional guidance. :)

# Load data

```{r data_load}
# Baby names (as seen in class)
baby_names <- read_csv("babynames.csv")


#install.packages("palmerpenguins")

library(palmerpenguins)

# ?palmerpenguins::penguins
# The code above will open the help page, where you can find a description
# of the penguins data set.

```

# filter()

## Question 1 

In 2018, what was the second-highest-ranked female baby name?

```{r}

```

## Question 2

In which year(s) was the name `"Eugene"` ranked among the top 20 names? What was its rank in each of those years?

```{r}

```

## Question 3

In which year(s) was the name `"Agnes"` ranked among the top 20 names?

```{r}

```

## Question 4

Create a bar plot of the top three female baby names since 2010. Interpret your plot.

```{r}

```

# arrange() 

## Question 1

Create a table showing the top-ranked male baby name for each year from 2010 to 2020. Arrange the table in descending order by year.

```{r}

```

## Question 2 

For 2020, create a vector containing the top 20 male baby names in alphabetical order.

```{r}

```

# summarise() and group_by()

## Question 1

Calculate the sample size and median bill length for each penguin species. Ignore missing values when calculating the median. Which species has the smallest observed median bill length?

```{r}

```

## Question 2

Create side-by-side boxplots of bill length for each penguin species. Make three observations based on the plot.

```{r}

```

## Question 3

Create boxplots of bill length for each penguin species, with the observations colour-coded by sex. Exclude observations with missing values for sex. Make two additional observations about the relationship between bill length and sex.

```{r}

```




# mutate()


## Question 1

Create a histogram of the ratio of flipper length to bill length for Adelie penguins using 15 bins. Change both the fill colour and outline colour of the histogram.

```{r}

```
## Question 2

How many Adelie penguins are in the data set?

Based on the histogram above, approximately what proportion of Adelie penguins have a flipper-to-bill length ratio less than or equal to 4.5?

a) 50%  
b) 25%  
c) 5%  
d) 2.5%

Which answer is closest?

```{r}

```

## Question 3

Calculate the exact proportion of Adelie penguins that have a flipper-to-bill length ratio less than or equal to 4.5.

```{r}

```